% -*- mode: context; -*-
\chapter{Periodicity of Trigonometrical Functions}
{\bf Definition of Periodic Functions:} A function $f(x)$ is said to be a periodic function if there
exists positive number $T$ independent of $x$ such that $f(x + T) = f(x),$ for every
$x\in$ domain $f$.

The least positive value of $T$ for which $f(x + T) = f(x)$, for every $x\in$ domain
$f$ is called the period of fundadmental period of $f(x)$.

{\bf Example 1:} Examine whether $\sin x$ is a periodic function or not. If yes, then find the period.

{\bf Sol.} Given, $f(x) = \sin x$. Let $f(x + T) = f(x) \Rightarrow \sin(x + T) = \sin x$

$\Rightarrow x + T = n\pi + (-1)^nx$, where $n = 0, \pm1, \pm2, \ldots$

The positive values of $T$ independent of $x$ are given by $T = n\pi$, where $n = 2,
4, 6, \ldots$

The least positive value of $T = 2\pi$.

Thus, $\sin x$ is a periodic function with period $2\pi$.

{\bf Some Results}
\startitemize[n]
\item $\sin x, \cos x, sec x$ and $\csc x$ are periodic functions with period $2\pi$.

\item $\tan x$ and $\cot x$ are periodic function with period $\pi$.

\item $|\sin x|, |\cos x|, |\tan x|, |\cot x|, |\sec x|$ and $|\csc x|$ are periodic functions
  with period $\pi$.

\item $\sin^nx, \cos^nx, \sec^nx$ and $\csc^nx$ are periodic functions with period $2\pi$
  and $\pi$ according as $n$ is odd or even.

\item $\tan^nx$ and $\cot^nx$ are periodic functions with period $\pi$ irrespective of
  $n$ being odd or even.

\item If a circular funciton $f(x)$ is periodic function with period $T$, then $kf(ax + b)$
  is also a periodic function with period $\frac{T}{|a|}$.

\item If circular functions $f(x)$ and $g(x)$ are periodic functions with period $T_1$ and
  $T_2$ then $af(x) + bg(x)$ is a periodic function with period $T$, where $T =$
  L.C.M. of $T_1$ and $T_2$.
  \stopitemize

\section{Problems}
\startitemize[n, 1*broad]
\item Which of the following functions are periodic? Also, find the period if the function is periodic.
  \startitemize[a, 1*broad, columns]
\item $f(x) = 10\sin3x$
\item $f(x) = a\sin\lambda x + b\cos\lambda x$
\item $f(x) = \sin^3x$ iv. $f(x) = \cos x^2$
\item $f(x) = \sin\sqrt{x}$
\item $f(x) = \sqrt{\tan x}$
\item $f(x) = x - |x|$ where $|x|$ is integral part of $x$
\item $f(x) = x\cos x$
  \stopitemize

\item Which of the following functions are periodic? Also, find the period if the function is periodic and has
  fundamental period.
  \startitemize[a, 1*broad, columns]
\item $f(x) = 4\sin\left(3x + \frac{\pi}{4}\right)$
\item $f(x) = 3\cos\frac{x}{2} + 4\sin\frac{x}{2}$
\item $f(x) = \cot\frac{x}{2}$
\item $f(x) = \sin^2x$
\item $f(x) = \sin x^2$
\item $f(x) = \sin\frac{1}{x}$
\item $f(x) = 1 + \tan x$
\item $f(x) = [x]$
\item $f(x) = 5$
\item $f(x) = |\cos x|$
\item $f(x) = \sin^4x + \cos^4x$
\item $f(x) = x + \sin x$
\item $f(x) = \cos\sqrt{x}$
\item $f(x) = \tan^{-1}(\tan x)$
\item $f(x) = |\sin x| + |\cos x|$
\item $f(x) = \sin\frac{\pi x}{3} + \sin\frac{\pi x}{4}$
\item $f(x) = \sin\left(2\pi x + \frac{\pi}{3}\right) + 2\sin\left(3\pi x + \frac{\pi}{4}\right) + 3\sin\pi x$
\item $f(x) = \sin x + \cos\sqrt{x}$
\stopitemize
\item Show that the function $f(x) = 2\sin x + 3\cos 2x$ is a periodic function of period $2\pi$.

\item For each of the following functions, mention whether the function is periodic and if yes, mention the
   period:
\startitemize[a]
\item $f(x) = 2x - [2x]$, where $[ ]$ denotes the integral part, and
\item $g(x) = 1 + \frac{3}{2 - \sin^2x}$
\stopitemize
\item Find the period of the function $1 - \frac{1}{4}\sin^2\left(\frac{\pi}{3} - \frac{3x}{2}\right)$.
  \stopitemize
